---
title: "Reproducible Research - Week 2 Assignment: Movement Analysis"
author: "Tingyao Wu"
date: "Monday, February 09, 2015"
output: html_document
---
# Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

# Data 
The data for this assignment can be downloaded from the course web site:

- Dataset: Activity monitoring data [52K]

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

# Assignment

## Loading and preprocessing the data

This chunk is to load the dataset, and show the summary of the data
```{r loading}
library(knitr)
library(lubridate)
setwd("C://Work//Learning//Courses//Data Science 5 - Reproducible Research//Week 2//project")
data <- read.csv("activity.csv")
data$date <- ymd(data$date)
summary(data)
```
## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r }
total_steps_per_day <- tapply(data$steps, data$date, sum, na.rm=TRUE)
xname <- c("total number of steps taken per day")
hist(total_steps_per_day, main = paste("Histogram of",xname), xlab = xname)
mean_steps_per_day <- tapply(data$steps, data$date, mean, na.rm=TRUE)
median_steps_per_day <- tapply(data$steps, data$date, median, na.rm=TRUE)
print(mean_steps_per_day)
print(median_steps_per_day)

```
## What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r }
average_step_per_interval <- tapply(data$steps, data$interval, mean, na.rm=TRUE)
plot(names(average_step_per_interval), average_step_per_interval,type="l",xlab="5-minute interval",ylab = "Average Steps", main="Average number of steps taken for every 5-minute interval"); 
grid(24,5)
name <- names(average_step_per_interval)
t <- which.max(average_step_per_interval)
 print(paste("The", name[t], "5-minute interval, on average across all the days in the dataset, contains the maximum number of steps"))
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
2. A strategy for filling in all of the missing values in the dataset. 
        
        - we assign the average number of steps of the same interval to the corresponding missing values
        
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
        
        - Yes. They differ. The impact is that: (1) the days containing missing values in the first step now are moved from the bin 0-5000 to the bin 10000 - 15000 (2) the days containing missing values have the same values after filled by the average.

```{r }
#nrow_na: the number of rows with NAs
nrow_na <- nrow(data[is.na(data$steps),])

print(paste("the total number of missing values:", nrow_na))

#we use the average number of steps of the same interval to the corresponding missing values
#the dataset with the filled missing values are stored in newdata
idx_na <- which(is.na(data$steps))
nadata <- data[idx_na,]
newdata <- data

interval_list <- unique(nadata$interval)
for (i in (1:length(interval_list)))
        {
        nadata[nadata$interval == interval_list[i], 1] <- average_step_per_interval[as.character(interval_list[i])]
        }
newdata[idx_na, ] <- nadata

#re-compute the mean and median
mean_steps_per_day_mv_filled <- tapply(newdata$steps, newdata$date, mean)
median_steps_per_day_mv_filled <- tapply(newdata$steps, newdata$date, median)
print(mean_steps_per_day_mv_filled)
print(median_steps_per_day_mv_filled)
total_steps_per_day_mv_filled <- tapply(newdata$steps, newdata$date, sum)
xname <- c("total number of steps taken per day after missing values filled")
hist(total_steps_per_day_mv_filled, main = paste("Histogram of",xname), xlab = xname)

```

## Are there differences in activity patterns between weekdays and weekends?
1 Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

2 Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r }
library(lattice)
newdata$day <- "weekday"
day <- weekdays(newdata$date)
weekend_idx <- which(day == "Saturday" | day == "Sunday" )
newdata$day[weekend_idx] <- "weekend"

#calculate weekdays_ave_step_per_interval and weekend_ave_step_per_interval
data_weekdays <- newdata[newdata$day == "weekday", ]
data_weekend <- newdata[newdata$day == "weekend", ]
weekdays_ave_step_per_interval <- tapply(data_weekdays$steps, data_weekdays$interval, mean)
weekend_ave_step_per_interval <- tapply(data_weekend$steps, data_weekend$interval, mean)

ave_step_dataframe <- data.frame(ave_step= c(weekdays_ave_step_per_interval, weekend_ave_step_per_interval), interval = c(names(weekdays_ave_step_per_interval), names(weekend_ave_step_per_interval) ), day = c(rep("weekday",each=length(weekdays_ave_step_per_interval)), rep("weekend",each=length(weekend_ave_step_per_interval))) )
ave_step_dataframe$interval <- as.numeric(as.character(ave_step_dataframe$interval))

#here we make a plot
xyplot(ave_step ~ interval|day,data = ave_step_dataframe, type = "l", layout = c(1,2))
```